---
title: "Getting and Cleaning Data Course Project"
author: "Catarina Lenice Lopes Pedrosa"
date: "05/07/2020"
output: html_document
---

## Human Activity Recognition Using Smartphones Dataset

The data has info about 30 volunteers that performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, it was captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain.

For each record it is provided:
- Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration.
- Triaxial Angular velocity from the gyroscope. 
- A 561-feature vector with time and frequency domain variables. 
- Its activity label. 
- An identifier of the subject who carried out the experiment.

The dataset includes the following files:

- 'README.txt'

- 'features_info.txt': Shows information about the variables used on the feature vector.

- 'features.txt': List of all features.

- 'activity_labels.txt': Links the class labels with their activity name.

- 'train/X_train.txt': Training set.

- 'train/y_train.txt': Training labels.

- 'test/X_test.txt': Test set.

- 'test/y_test.txt': Test labels.

## Features

The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

tBodyAcc-XYZ
tGravityAcc-XYZ
tBodyAccJerk-XYZ
tBodyGyro-XYZ
tBodyGyroJerk-XYZ
tBodyAccMag
tGravityAccMag
tBodyAccJerkMag
tBodyGyroMag
tBodyGyroJerkMag
fBodyAcc-XYZ
fBodyAccJerk-XYZ
fBodyGyro-XYZ
fBodyAccMag
fBodyAccJerkMag
fBodyGyroMag
fBodyGyroJerkMag

The set of variables that were estimated from these signals are: 

mean(): Mean value
std(): Standard deviation
mad(): Median absolute deviation 
max(): Largest value in array
min(): Smallest value in array
sma(): Signal magnitude area
energy(): Energy measure. Sum of the squares divided by the number of values. 
iqr(): Interquartile range 
entropy(): Signal entropy
arCoeff(): Autorregresion coefficients with Burg order equal to 4
correlation(): correlation coefficient between two signals
maxInds(): index of the frequency component with largest magnitude
meanFreq(): Weighted average of the frequency components to obtain a mean frequency
skewness(): skewness of the frequency domain signal 
kurtosis(): kurtosis of the frequency domain signal 
bandsEnergy(): Energy of a frequency interval within the 64 bins of the FFT of each window.
angle(): Angle between to vectors.

## Dowloading and unziping the data

```{r}
download.file("https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip", "project")

unzip("project")
```

## Reading files into R

```{r}
setwd("/home/catarina/Área de Trabalho/FinalProject/UCI HAR Dataset")

features <- read.table("features.txt")
activity <- read.table("activity_labels.txt", stringsAsFactors = FALSE)

setwd("/home/catarina/Área de Trabalho/FinalProject/UCI HAR Dataset/test")

Y_test <- read.table("y_test.txt")
X_test <- read.table("X_test.txt")
Subject_test <- read.table("subject_test.txt")

setwd("/home/catarina/Área de Trabalho/FinalProject/UCI HAR Dataset/train")

Y_train <- read.table("y_train.txt")
X_train <- read.table("X_train.txt")
Subject_train <- read.table("subject_train.txt")
```

## Binding the datasets 

Binding the test and train datas and extracting only the measurements on the mean and standard deviation for each measurement

```{r}

setwd("/home/catarina/Área de Trabalho/FinalProject")

X_full <- rbind(X_test, X_train)
Y_full <- rbind(Y_test, Y_train)
Subject_full <- rbind(Subject_test, Subject_train)

columns_measurements <- c(grep("std()", features$V2), grep("mean()", features$V2))

X_full <- X_full[,columns_measurements]

```

## Renaming columns and joinning X and Y

Renaming columns to descriptive names according to the data into feature file, and binding the X and Y data.

```{r}

names(X_full) <- features$V2[columns_measurements]

names(Y_full)[1] <- "activities"

names(Subject_full)[1] <- "subject"

df <- cbind(Y_full,Subject_full, X_full)

```

## Naming activities

Using the descriptive activity names in the activitys file to name the activities in the data set.

```{r}
df <- df %>% 
  mutate(activities = replace(activities, activities == 1, activity$V2[1])) %>% 
  mutate(activities = replace(activities, activities == 2, activity$V2[2])) %>% 
  mutate(activities = replace(activities, activities == 3, activity$V2[3])) %>% 
  mutate(activities = replace(activities, activities == 4, activity$V2[4])) %>% 
  mutate(activities = replace(activities, activities == 5, activity$V2[5])) %>% 
  mutate(activities = replace(activities, activities == 6, activity$V2[6]))
```

## Creating new dataset

Creating new dataframe by grouping the data by activities and getting the average of each variable.

```{r}
df_by_activity <- df %>% 
  group_by(subject,activities) %>% 
  summarise_all(list(mean))
```

## Writing the data into txt file

```{r}
write.table(df_by_activity, "df_by_activity", row.name=FALSE)
```